A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-11-01 thesis/thesis.tex
add execution example

To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh()} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to the $n$. We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
We sketch the argument here.

@@

\Tree [.{\begin{tabular}{||c|c|c||}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline (2,1) & (5,0) & (4,2) & (3,4) & (6,3) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption{\label{fig:block} Using blocks to represent operations. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively. For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child. (5,3) is superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

@@

\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
\State \tt{blocks[\head]= B} 
\State \tt{\head+=1} \label{appendEnd} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

@@

>>>>

To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to the $n$. We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
We sketch the argument here.

@@

\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline (2,1) & (5,0) & (4,2) & (3,4) & (6,3) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
\end{small}
\end{tikzpicture}
\end{center}
\caption{\label{fig:block} Using blocks to represent operations. Blocks between two lines $||$ are propagated together to the parent. Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively. For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child. The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child. (5,3) is superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
\end{figure}

@@

\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
\State \tt{blocks[\head]= B} \label{appendLeaf}
\State \tt{\head+=1} \label{appendEnd} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

@@

\pagebreak
\section{Appendix}
\subsection{Example Execution of \nf{Propagate}}

Here we want to demonstrate how four processes help each other to  propagate their operations up to the root. We show how the steps of each \nf{Propagate} works as the time goes.

In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$. Line \ref{appendLeaf} appends block $B$ to the leaf.

Then as in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ and propagates both $P1$ and $P2$ operations. $P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$. $P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after the $P2$'s successful \nf{CAS} that put $ENQ_1,ENQ_2$. So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s \ref{readHead} is after $P2$'s \ref{cas}).

Since the block $P1$ creates is empty it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}. $P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it reaches $DEQ_4$ to the \nf{root} even before $P2$ puts its block to $n5$.

Finally $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root. After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ to the \nf{root}. $P1$ fails to do \nf{Refresh} on the \nf{root} two times, but for sure its operation is propagated after the second failed \nf{Refresh}. See Figure \ref{fig::exec4}

All the fields of some blocks from the Figure \ref{fig::exec4} are filled in Figure \ref{fig::exec45}.

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
  \caption{$P1,P2,P4$ append operations. \label{fig::exec1}}
\end{figure}

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
  \caption{$P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
\end{figure}

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
  \caption{$P4$ puts its block to the \nf{root} before $P2$ appending its block to $n5$.\label{fig::exec3}}
\end{figure}

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
  \caption{$P3$ suceeds to the \nf{CAS} to the \nf{root}. $P1$'s work is done after two failed \nf{CAS}es. \label{fig::exec4}}
\end{figure}

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
\end{figure}
